For Induction-Induction, Induction is Enough

### 2019-11-11 paper/paper.tex
example sigs and notation

\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)}

Most mutual inductive types can be reduced to indexed inductive types
where the index gives the information which sort we meant. An example
is the predicates $\mathsf{isEven}$ and $\mathsf{isOdd}$ defined
mutually by the following constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,n \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,n
\end{alignat*}
These can be reduced to the following single inductive family.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,n \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,n
\end{alignat*}

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the James Chapman style
\cite{chapman09eatitself} syntax for type theory. A minimal use case
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$
constructor of $\Con$ refers to $\Ty$ and the $\Ty$-constructor
$\mathsf{U}$ refers to $\Con$, hence the two sorts have to be defined
at the same time. Moreover, $\Ty$ is indexed over $\Con$, which
precludes the above reduction which worked for
$\mathsf{isEven}$--$\mathsf{isOdd}$ as we would get a type indexed
over itself. Another unique feature of IITs (which also shows up in
higher inductive types \cite{HoTTbook}) is that later constructors can
refer to previous constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for an IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
given the following components,
\begin{alignat*}{5}
  & Con^M && : \Con\ra\Set \\
  & Ty^M && : Con^M\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & empty^M && : Con^M\,\mathsf{empty} \\
  & ext^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,A\ra Con^M\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,(\mathsf{U}\,\Gamma) \\
  & El^M && : (\Gamma^M:Con^M\,\Gamma)\ra\,Ty^M\,(ext^M\,\Gamma^M\,(U^M\,\Gamma^M))\,(\mathsf{El}\,\Gamma)
%  & Pi^M && : (\Gamma^M:Con^M\,\Gamma)(A^M:Ty^M\,\Gamma^M\,A)\ra Ty^M\,(ext^M\,\Gamma^M\,A^M)\ra\,Ty^M\,\Gamma^M\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^M\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = empty^M \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^M\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^M\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

As opposed to mutual inductive types, reducing IITs to inductive types
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction
which only supports a simpler, non recursive-recursive elimination
principle (\todo{check}). Hugunin \cite{jasper} reduced an example IIT
to inductive types without UIP, but he also only constructed the
simpler non recursive-recursive eliminator.

@@

We work in intensional Martin-Löf type theory, equipped with the following features:
\begin{itemize}
  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
    non-dependent. We sometimes omit function arguments, by implicitly
    generalising over variables.
  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
        when we want to leave the type of the first projection implicit. We use $A
        \times B$ for non-dependent pairs.
  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
    this paper, we use extensional notation, and do not write explicit
    transports over equalities. However, we occasionally indicate by
    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
    $e_1$,\dots,$e_n$.
  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.

@@

We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
for inductive parameters, we have $\app$ as a categorical version of function
application. We recover the usual type-theoretic application as follows:
\begin{alignat*}{5}
  & \blank\appm\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & t \appm u := \app\,t\,[\id\,u]
\end{alignat*}
We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
$p$ composed with itself $n$ times. However, for informal examples we instead
use a nameful notation, where we introduce names in context extension and in the
$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,(A : \U)\,\ext\,(a :
\El\,A)$ in a signature for pointed sets.

@@

@@

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, taking the signature of natural numbers:
\[
  \Gamma := \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

>>>>

Many mutual inductive types can be reduced to indexed inductive types, where the
index disambiguates different sorts. An example is the predicates
$\mathsf{isEven}$ and $\mathsf{isOdd}$ defined mutually by the following
constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,n \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,n
\end{alignat*}
These can be reduced to the following single inductive family.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,n \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,n
\end{alignat*}

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimal use case
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$
constructor of $\Con$ refers to $\Ty$ and the $\Ty$-constructor
$\mathsf{U}$ refers to $\Con$, hence the two sorts have to be defined
at the same time. Moreover, $\Ty$ is indexed over $\Con$, which
precludes the above reduction which worked for
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed
over itself. Another unique feature of IITs (which also shows up in
higher inductive types \cite{HoTTbook}) is that later constructors can
refer to previous constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions, where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
given the following components,
\begin{alignat*}{5}
  & Con^M && : \Con\ra\Set \\
  & Ty^M && : Con^M\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & empty^M && : Con^M\,\mathsf{empty} \\
  & ext^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,A\ra Con^M\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,(\mathsf{U}\,\Gamma) \\
  & El^M && : (\Gamma^M:Con^M\,\Gamma)\ra\,Ty^M\,(ext^M\,\Gamma^M\,(U^M\,\Gamma^M))\,(\mathsf{El}\,\Gamma)
%  & Pi^M && : (\Gamma^M:Con^M\,\Gamma)(A^M:Ty^M\,\Gamma^M\,A)\ra Ty^M\,(ext^M\,\Gamma^M\,A^M)\ra\,Ty^M\,\Gamma^M\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^M\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = empty^M \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^M\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^M\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

As opposed to mutual inductive types, reducing IITs to inductive types is an
open problem. Forsberg \cite{forsberg-phd} showed a reduction which only
supports a simpler, non-recursive-recursive elimination principle
(\todo{check}). Hugunin \cite{jasper} reduced several IITs to inductive types,
working inside a UIP-free cubical type theory, but he also only constructed the
simpler non-recursive-recursive eliminator.

@@

We work in intensional Martin-Löf type theory, equipped with the following features:
\begin{itemize}
  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
    non-dependent. We sometimes omit function arguments, by implicitly
    generalising over variables.
  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
        when we want to leave the type of the first projection implicit. We use $A
        \times B$ for non-dependent pairs.
  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
    this paper, we informally use extensional notation, and do not write
    explicit transports over equalities. However, we occasionally indicate by
    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
    $e_1$,\dots,$e_n$.
  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.

@@

We define $\sigma^\uparrow$ as $(\sigma\circ\p,\q)$. In the $\Pi$ function type
for inductive parameters, we have $\app$ as a categorical version of function
application. We recover the usual type-theoretic application as follows:
\begin{alignat*}{5}
  & \blank\oldapp\blank : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & t \oldapp u := \app\,t\,[\id\,u]
\end{alignat*}
We also abbreviate non-dependent $\Pi$ functions by writing $\Pi\,a\,(B[p])$ as $a\ra B$.

We can also recover de Bruijn indices by having $n := q[p^n]$, where $p^n$ is
$p$ composed with itself $n$ times. However, for informal examples we instead
use a nameful notation, where we introduce names in context extension and in the
$\Pi$ function domain, for example as $\boldsymbol{\cdot}\,\ext\,A : \U\,\ext\,a :
\El\,A$ in a signature for pointed sets.

@@

Natural numbers:
\[
\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
Lists with elements of a given $A : \Set$ type. Here we use the $\Pim$ function
to include an external type into the signature.
\[
\boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,
      cons : \Pim\,A\,(\lambda \_.\, L\ra L)
\]
The inductive-inductive Con-Ty example from Section \ref{sec:intro}:
\begin{alignat*}{3}
  & \boldsymbol{\cdot} \,\ext\,Con : \U\,\ext\,Ty : Con\ra\U\,\ext\,empty : \El\,Con\,\ext\,ext : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \ra Con) \\
  & \hspace{0.5em}\,\ext\, U : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))
    \,\ext\,El : \Pi\,(\Gamma : Con)(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp U)))
\end{alignat*}

  %% & \Ty && : \Con\ra\Set \\
  %% & \mathsf{empty} && : \Con \\
  %% & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  %% & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
  %% & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))

@@

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, considering the signature for natural
numbers
\[
  \Gamma := \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

### 2019-11-11 paper/universaliit.tex
example sigs and notation

  UIP\cite{hugunin2019constructing}, and demonstrated that well-formedness
  predicates used in syntactic models can be subtly incompatible with UIP. Also,
  while Hugunin does not use UIP, he only shows a weak version of dependent
  elimination for the constructed IITs. Hence, the question whether IITs are
  reducible to inductive types in a UIP-free setting remains open.

>>>>

  UIP\cite{hugunin2019constructing} in cubical type theory, and demonstrated
  that well-formedness predicates used in syntactic models can subtly break
  in that setting. Also, while Hugunin does not use UIP, he only shows a
  weak version of dependent elimination for the constructed IITs. Hence, the
  question whether IITs are reducible to inductive types in a UIP-free setting
  remains open.
